---
title: "euclide_gene_pca_loadings"
author: "Natasha Howe"
date: "`r Sys.Date()`"
output: html_document
---

### PCAngsd Loadings for Gene Regions

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries-metadata, warning=F, message=F}
library(tidyverse)
library(cowplot)
```

Slurm script with the following code was run for lrrc9 and esr1. I ran it with `--e 2` and `--e 10`

```{bash, eval=F}
module load bio/pcangsd/1.36.4

path=/home/nhowe/runtiming/anvil
prefix=euclide

# Region Specific naming, conducted separately for lrrc9
gene=esr1
chrom=NC_068428.1
firstpos=23264160
lastpos=23645630

nEigen=2

# call in whole genome beagle file
BEAGLE=${path}/gls/${prefix}_wholegenome_minInd0.3.beagle.gz

# cut to regions of interest
zgrep -E "${chrom}" ${BEAGLE} | awk -v s=$firstpos -v e=$lastpos -F'[\t_]' '$3 >= s && $3 <= e' | gzip > ${path}/gls/${prefix}_${chrom}_s${firstpos}_e${lastpos}_minInd0.3_${gene}.beagle.gz

echo "Create pca loading weights file with 2 pcs"
pcangsd --threads 2 \
  -b ${path}/gls/${prefix}_${chrom}_s${firstpos}_e${lastpos}_minInd0.3_${gene}.beagle.gz \
  -o ${path}/pca/${prefix}_${chrom}_s${firstpos}_e${lastpos}_minInd0.3_${gene}_e2 \
  --sites-save --eig ${nEigen} --snp-weights --iter 500 --pcadapt

# Create snp list        
zcat ${path}/gls/${prefix}_${chrom}_s${firstpos}_e${lastpos}_minInd0.3_${gene}.beagle.gz | cut -f1 | awk -F"_" '{print $1"_"$2"\t"$3}' > ${path}/gls/${prefix}_${chrom}_s${firstpos}_e${lastpos}_minInd0.3_${gene}.snp_list
```

### ESR1

```{r}
path= "../../../../Salmon_runtiming/2024_sockeye/results"
outdir="../../../../Salmon_runtiming/2024_sockeye/figures/pca/"
filename="euclide_NC_068428.1_s23264160_e23645630_minInd0.3_esr1"
Neigen=2

# PC weights
weight_tmp <- read_delim(paste0(path,"/pca/",filename,"_e",Neigen,".weights"), delim=" ",
                         col_names = c("pc1","pc2"), show_col_types = F) %>% 
  as_tibble()

# output sites file from PCAngsd - not as expected
# should be 1 for keep and 0 for remove
sites_file <- read_tsv(str_c(paste0(path,"/pca/",filename,"_e",Neigen,".sites")), 
                       col_names="keep", show_col_types = F)

# snp-list from beagle file
snp_list <- read_tsv(str_c(paste0(path,"/gls/",filename,".snp_list")), 
                     col_names = c("lg","pos"),
                     show_col_types = F) %>% slice(-1)

# weight table from snp list and kept sites  
weight_table <- sites_file %>%
  bind_cols(snp_list) %>%
  filter(keep==1) %>%
  dplyr::select(lg, pos) %>%
  bind_cols(weight_tmp) %>%
  pivot_longer(cols = 3:(ncol(weight_tmp)+2), names_to = "pc", values_to = "weight")

weight_table %>%
  filter(pc=="pc1" | pc=="pc2") %>%
  mutate(pos = as.numeric(pos)) %>%
  ggplot() +
  geom_point(aes(x=pos/10^6, y=abs(weight)), size=1) +
  scale_x_continuous(breaks=seq(0, 100, 0.05)) +  
  facet_grid(pc~., scales='free_x', space='free_x') +
  labs(x = "Position (Mb)",
       title = "Sockeye: ESR1") +
  theme(panel.spacing = unit(0.1, 'lines'),
        axis.title.x=element_text(),
        axis.text=element_text(size=6),
        legend.position='none',
        text = element_text(size=10),
        axis.text.x = element_text(angle = 45))
```

##### ESR1 Test on Broader region

Just a check on the **unused** wider boundary for esr1

```{r}
path= "../../../../Salmon_runtiming/2024_sockeye/results"
outdir="../../../../Salmon_runtiming/2024_sockeye/figures/pca/"
filename="euclide_NC_068428.1_s23141735_e23906114_minInd0.3_esr1"
Neigen=2

# PC weights
weight_tmp <- read_delim(paste0(path,"/pca/",filename,"_e",Neigen,".weights"), delim=" ",
                         col_names = c("pc1","pc2"), show_col_types = F) %>% 
  as_tibble()

# output sites file from PCAngsd - not as expected
# should be 1 for keep and 0 for remove
sites_file <- read_tsv(str_c(paste0(path,"/pca/",filename,"_e",Neigen,".sites")), 
                       col_names="keep", show_col_types = F)

# snp-list from beagle file
snp_list <- read_tsv(str_c(paste0(path,"/gls/",filename,".snp_list")), 
                     col_names = c("lg","pos"),
                     show_col_types = F) %>% slice(-1)

# weight table from snp list and kept sites  
weight_table <- sites_file %>%
  bind_cols(snp_list) %>%
  filter(keep==1) %>%
  dplyr::select(lg, pos) %>%
  bind_cols(weight_tmp) %>%
  pivot_longer(cols = 3:(ncol(weight_tmp)+2), names_to = "pc", values_to = "weight") 

weight_table %>%
  mutate(pos = as.numeric(pos)) %>%
  ggplot() +
  geom_point(aes(x=pos/10^6, y=abs(weight)), size=1) +
  scale_x_continuous(breaks=seq(0, 100, 0.05)) +  
  facet_grid(pc~., scales='free_x', space='free_x') +
  labs(x = "Position (Mb)",
       title = "Sockeye: ESR1 - Test on Wider Boundary") +
  theme(panel.spacing = unit(0.1, 'lines'),
        axis.title.x=element_text(),
        axis.text=element_text(size=6),
        legend.position='none',
        text = element_text(size=10),
        axis.text.x = element_text(angle = 45))
```

### LRRC9

```{r}
filename="euclide_NC_068455.1_s28128954_e28169980_minInd0.3_lrrc9"
Neigen=2

# PC weights
weight_tmp <- read_delim(paste0(path,"/pca/",filename,"_e",Neigen,".weights"), delim=" ",
                         col_names = c("pc1","pc2"), show_col_types = F) %>% 
  as_tibble()

# output sites file from PCAngsd - not as expected
# should be 1 for keep and 0 for remove
sites_file <- read_tsv(str_c(paste0(path,"/pca/",filename,"_e",Neigen,".sites")), 
                       col_names="keep", show_col_types = F)

# snp-list from beagle file
snp_list <- read_tsv(str_c(paste0(path,"/gls/",filename,".snp_list")), 
                     col_names = c("lg","pos"),
                     show_col_types = F) %>% slice(-1)

# weight table from snp list and kept sites  
weight_table <- sites_file %>%
  bind_cols(snp_list) %>%
  filter(keep==1) %>%
  dplyr::select(lg, pos) %>%
  bind_cols(weight_tmp) %>%
  pivot_longer(cols = 3:(ncol(weight_tmp)+2), names_to = "pc", values_to = "weight") # %>% left_join(chr_table, by="lg")

weight_table %>%
  mutate(pos = as.numeric(pos)) %>%
  ggplot() +
  geom_point(aes(x=pos/10^6, y=abs(weight)), size=1) +
  scale_x_continuous(breaks=seq(0, 100, 0.01)) +  
  facet_grid(pc~., scales='free_x', space='free_x') +
  labs(x = "Position (Mb)",
       title = "Sockeye: LRRC9") +
  theme(panel.spacing = unit(0.1, 'lines'),
        axis.title.x=element_text(),
        axis.text=element_text(size=6),
        legend.position='none',
        text = element_text(size=10),
        axis.text.x = element_text(angle = 45))

```

### Selection

DID NOT TOUCH THE REMAINDER OF THIS FOR SOCKEYE!

PCAngsd has two implemented selection scans based on (i) FASTPCA and (ii) pcadapt. The associated paper compares the methods with high and low coverage data ([Meisner et al. 2021](https://pmc.ncbi.nlm.nih.gov/articles/PMC8480091/#Sec2)).

Also, this is code for the `--selection` flag for FASTPCA, but I ran the `--pcadapt` for PCADAPT. And this is for whole genome

```{r}
############### nicolas lou rmd
selection_tmp <- read_delim(paste0(path,'/pca/',filename,"_e",Neigen,".selection"), 
                            delim=" ", col_names = F, show_col_types = F) %>% 
  as_tibble()

# combine selection output with sites
selection_table <- sites_file %>%
  bind_cols(snp_list) %>%
  filter(keep==1) %>%
  dplyr::select(lg, pos) %>%
  bind_cols(selection_tmp) %>%
  pivot_longer(cols = 3:(ncol(selection_tmp)+2), names_to = "pc", values_to = "d") %>%
  mutate(p=pchisq(d, 1, lower.tail=FALSE)) %>%
  left_join(chr_table, by="lg") %>%
  group_by(pc) %>%
  mutate(bonferroni=p.adjust(p, method="bonferroni"),
         fdr=p.adjust(p, method="fdr")) %>%
  ungroup()

# Plot for PC1 and PC2
selection_table %>%
  ggplot(aes(x=as.numeric(pos)/10^6, y=-log10(p), color=as_factor(chr))) +
  geom_point(size=0.1) +
  geom_point(data=selection_table %>% filter(fdr<0.001), color="red", size=0.2) +
  # geom_point(data=selection_table %>% slice_min(p, n=round(nrow(selection_tmp)/10000)),
  #            color="red", size=0.2) +
  scale_color_manual(values = rep(c('black', 'grey50'), 50)) +
  scale_x_continuous(breaks=seq(10, 90, 10)) +  
  facet_grid(pc~chr, scales='free_x', space='free_x') +
    labs(x = "Position (Mb)",
       title = "Sockeye: PCAngsd Selection Scan") +
  theme_cowplot() +
  theme(panel.spacing = unit(0.1, 'lines'),
        axis.title.x=element_text(),
        axis.text=element_text(size=6),
        legend.position='none',
        text = element_text(size=10),
        axis.text.x = element_text(angle = 45))

ggsave(paste0(outdir,"euclide_pcangsd_selection_scan.jpg"), height = 10, width = 20)
```

```{r}
selection_table %>% filter(fdr<0.001) %>%
   count(chr)
```
